---
title: "BDA_020: Pointers on Joint Distributions"
output: html_notebook
---

A joint distribution describes the probability of multiple events occuring. The ones that Blitzstein draws attention to in his book are the multinomial, multivariate Normal, and cauchy distributions.

The multinomial distribution describes the probability of observing $n_j$ of each category in $N$ draws, such that $\sum_{j = 1}^P n_j = N$. 

```{r}
x <- c(2, 0, 3) # Probability of 2 in first cat., 0 in second, 3 in third
n <- 5          # Total number of trials
p <- c(1/3, 1/3, 1/3) # Probability of each category

dmultinom(x, n, p) # Probability density at (2, 0, 3)

rmultinom(1, size = n, prob = p) # A random draw from the Mult(n, p) distribution
```

To work with the multivariate Normal distribution, we need to use the $\texttt{mvtnorm}$ package. 
```{r}
library(mvtnorm)
library(plot3D)
mu <- c(0, 0)
rho <- 0
covmat <- matrix(c(1, rho, rho, 1), nrow = 2, ncol = 2)
n_bins = 25
x <- seq(-3, 3, length.out = n_bins)
y <- seq(-3, 3, length.out = n_bins)
r <- matrix(dmvnorm(expand.grid(x, y), mu, covmat), nrow = n_bins)
ribbon3D(z = r, border=NA, phi = 40, curtain = FALSE, facets = TRUE, col = ramp.col(c("orange", "red")), resfac = 1, colkey = FALSE)
```


All right, that's enough farting around with 3D plots. The cauchy distribution is also available - this isn't a joint distribution in itself, since a Cauchy r.v. $T = \frac{X}{Y}$ where $X, Y \sim N(0, 1)$.
```{r}
hist(rcauchy())
```