---
title: "BDA_021: The Metropolis Algorithm"
output: html_notebook
---

We have a chain of nodes. Each node has a unique mass. These masses sum to 1. It's possible to move between adjacent nodes. At each step, we flip a coin to determine in which direction to move. Once a direction is chosen, we move to that node with probability $\text{min}(\frac{m_{\text{adj}}\}{m_{\text{current}}}, 1)$. We do this many many times.

```{r, echo = FALSE}
nodes <- c(0.1, 0.1, 0.1, 0.7)
current_node <- 2
n_steps <- 10000
chain <- matrix(current_node, nrow = n_steps, ncol = 1)
for (i in seq(2, n_steps)) {
  proposed_node <- ifelse((runif(1) > 0.5),
                           current_node + 1, current_node - 1)
  proposed_node <- ifelse(proposed_node == 5, 1, proposed_node)
  proposed_node <- ifelse(proposed_node == 0, 4, proposed_node)
  current_node  <- ifelse(rbinom(1, 1, nodes[proposed_node]) == 1, 
                          proposed_node, current_node)
  chain[i] <- current_node
}
{plot(seq(n_steps), chain, type = 'l', pch = 20, log = 'x')
lines(seq(n_steps), chain, type = 'p', pch = 20)}
```

Off-screen, we showed that the posterior is the stationary distribution of a Markov chain for which the transitions occur according to the Metropolis algorithm.

Kruschke asked us to install BRugs, so we did. He provides a demo model:
```{r}

library(rstan) # Load Stan API

fileNameRoot="Stan-BernBeta-Script-"

# Specify model:
modelString = "
  data {
    int<lower=0> N ; # Number of trials
    int y[N] ;       # Array of trial results
  }
  parameters {
    real<lower=0,upper=1> theta ; # Success probability
  }
  model {
    theta ~ beta(1,1) ;     # Prior
    y ~ bernoulli(theta) ;  # Likelihood
  }
" # close quote for modelString

stanDso <- stan_model( model_code=modelString ) # Constructs a model in Stan, then returns an object to interact with model

# Specify data:
N = 50 ; z = 10
y = c(rep(1,z), rep(0,N-z)) # z 1s, N-z 0s
dataList = list(
  y = y,
  N = N 
)

# Generate posterior sample:
stanFit <- sampling( object=stanDso , 
                     data = dataList , 
                     chains = 3 ,
                     iter = 1000 , 
                     warmup = 200 , 
                     thin = 1 )

traceplot(stanFit, pars=c("theta"))
plot(stanFit, pars=c("theta"))

{openGraph()
mcmcCoda = mcmc.list( lapply( 1:ncol(stanFit) , 
                              function(x) { mcmc(as.array(stanFit)[,x,]) } ) )
diagMCMC( mcmcCoda , parName=c("theta") )
saveGraph(file=paste0(fileNameRoot,"Diag"),type="pdf")
}
```